Intersection Type Algebras

### 2025-01-17 UTA/notes.tex
uta

\newcommand{\types}{\mathbb{T}}

@@

>>>>

\input{lambdamacs}

% \newcommand{\types}{\mathbb{T}}
\newcommand{\mcS}{\mathcal{S}}
\newcommand{\mcF}{\mathcal{F}}
\newcommand{\semof}[1]{{[\![{#1}]\!]}}
\newcommand{\ita}{\mathpzc{Ita}}
\newcommand{\msl}{\mathpzc{Msl}}

@@

\section*{01.17}
TODO:
\begin{itemize}
  \item Need to start writing the paper!
  \item Need a pretty-printer/helper functions for using the Agda code.
  \item Need to refactor/cleanup the code
  \item Work on the conjecture
  \begin{enumerate}
    \item One direction should be easy
    \item Is the other direction even true?
  \end{enumerate}
  \item Generalizing the code to preorders/intersection types
  \item Verify correctness of the code: prove it computes the most general unifier
  equivalent to the starting list of constraints
  \item Implementing the decision algorithm for the generated recursive type algebra
  \item How to implement/define the ``free'' ita/meet semilattices in type theory
  --- related to: How to define strict order between types.
  \item Implement/verify the decider for the subtyping relation between trees.
\end{itemize}

\section{Notes from jan. 2025}
Intersection type algebras are the simultaneous free algebra for the
monoid functor (whose free algebras are list algebras) and
icmonoid functor (whose free algebras are free semilattices).

Specifically, given a type $A$, let $\mathcal{P}_f(A)$ denote the final
powerset of $A$.  Define
\begin{align}
  \label{e:msl}
  \msl(A) &= \mathcal{P}_f(A) \\
  \label{e:ita}
  \ita(A) &= \mu X. A \times \msl(X)^*
\end{align}

That is, for any type $A$, $\ita(A)$ is the initial type satisfying
$\ita(A) = \msl(\ita(A))^* \times A$.

The distributivity is implicit in this representation, since the
type $\msl(\ita(A))$ forms an intersection type algebra in the sense of
BCD, which involves two binary operations, $\to$ and $\cap$.
Specifically, given $S_1,S_2 \sse_f (\ita(A)$, we
define
\begin{align*}
  (S_1 \cap S_2) &= S_1 \cup S_2 \\
  (S_1 \to S_2) &= \setof{((s:ts),a) \mid s \in S_1, (ts,a) \in S_2}
\end{align*}
This translation validates the distributivity rule, since
\begin{align*}
  S_1 \to (S_2 \cap S_3)
  &= S_1 \to (S_2 \cup S_3) \\
  &= \setof{(s:ts,a) \mid s \in S_1, (ts,a) \in S_2 \cup S_3}\\
  &= \setof{(s:ts,a) \mid s \in S_1, (ts,a) \in S_2}
  \cup \setof{(s:ts,a) \mid s \in S_1, (ts,a) \in S_3}\\
  &= (S_1 \to S_2) \cup (S_1 \to S_3) \\
  &= (S_1 \to S_2) \cap (S_1 \to S_3)
\end{align*}

With the monadic unit $\eta_A = \lam a{:}A. (\emptyset,a): A \to \msl(\ita(A))$,
and the initiality of the algebra of BCD types $\types_{\to,\cap}(A)$,
we obtain a map $dnf_A : \types_{\to,\cap)(A) \to \msl(\ita(A))}$.

The converse map $bcd_A : \msl(\ita(A)) \to \types_{\to,\cap}(A)$
is given recursively by
\begin{align*}
  bcd_A(\setof{T_1,\dots,T_m}) &= bcd'_A(T_1) \cap (bcd'_A(T_2) \cap
    \cdots (bcd'_A(T_{m-1}) \cap bcd'_A(T_m)) \cdots)\\
  bcd'(A)([S_1,\dots,S_n],a) &= bcd_A(S_1) \to (bcd_A(S_2) \to \cdots (bcd_A(S_n)\to a)
  \cdots)
\end{align*}

It is easy to verify by induction that $bcd_A$ is injective.
It therefore splits the map $dnf_A$, computing the canonical representative
for the BCD-equivalence class of every type in $\types_{\to,\times}(A)$ .

\emph{Need to choose an ordering in representation of subsets!}
